### 2014-01-11 thesis-proposal/background.tex
Final Sec. 1

% !TEX root = omar-thesis-proposal.tex
\subsubsection{Background: Active Libraries}\label{alibs}
The term \emph{active libraries} was first introduced by Veldhuizen et al. \cite{atthesis, atpaper} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''. The authors went on to suggest a number of concrete reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler. Our definition, given in Section \ref{motivation}, differs  from theirs, partly for this reason).

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{blitzpp}. Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about \cite{Robison:2001:IEC:376656.376751}. %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

More powerful compile-time \emph{term rewriting mechanisms} integrated into some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending the language with new abstractions. These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety. Compile-time macros, such as those in MetaML \cite{metaml}, Template Haskell \cite{template_haskell} and Scala \cite{scala-macros}, can take full control over all the code that they enclose, so outer macros can neglect to invoke or  manipulate the output of inner macros, sometimes inadvertantly. Once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problem described in Section \ref{motivation}). Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that can handle them.

Some rewriting systems go beyond the block scoping of macros. Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{xroma}. Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{haskell-rewrite}. In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros. But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently. It is also difficult to guarantee that such logic is correct and difficult for users to reason about code when simply importing a library can change the semantics of the program in a non-local manner.
% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html

Another example of an active library approach to extensibility is SugarJ \cite{sugarj} and other languages in the Sugar* family \cite{sugarstar}, like SugarHaskell. These languages permit libraries to extend the base syntax of the core language in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program. These extensions are also not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same standard notations but back them with differing implementations. 
And again, it can be difficult for users to predict what type of data an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.

>>>>

% !TEX root = omar-thesis-proposal.tex
\subsubsection{Background}\label{alibs}
The term \emph{active libraries} was first introduced by Veldhuizen et al. \cite{activelibraries, active-libraries-thesis} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''. The authors suggested a number of reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}. Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}). %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending a language with new abstractions. These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety. For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all the code that they enclose. This can be problematic, however, as outer macros can interfere with the functionality of inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problems described in Section \ref{motivation}). Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.

Some term rewriting systems replace the block scoping of macros with pattern-based dispatch. Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}. Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}. In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros. But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently. It is also difficult to guarantee that such logic is correct and difficult to reason about code when importing a library can change the semantics of the program in a non-local manner.
% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html

Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}. These languages permit libraries to extend the base syntax of the core language in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program. These extensions are also not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same standard notations but back them with differing implementations. 
And again, it can be difficult for users to predict what type of data an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.

### 2014-01-11 thesis-proposal/problem.tex
Final Sec. 1

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new language and its associated tools altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well.
But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about  interoperability. F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support advanced features like polymorphic variants, modules or functors \cite{ocaml-manual}.
%\end{itemize}

\subsection{Language-Integrated Approaches}
We argue that, due to these compatibility problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary. More specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{elaboration} and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{active-libraries}. Features implemented within active libraries are imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, avoiding the compatibility problem.

We must proceed with caution, however: critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable. 
%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries. Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics and lead to subtle problems at link-time. For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program. These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.

>>>>

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well.
But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about  interoperability. F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support features like polymorphic variants, modules or functors \cite{ocaml-manual}.
%\end{itemize}

\subsection{Language-Integrated Approaches}
We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{elaboration} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}. Features implemented within active libraries can be imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, seemingly avoiding the client compatibility problem.

We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable. 
%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries. Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time. For example, if two active libraries introduce the same syntactic form but back it with differing (valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit. These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
